# pre-class


Make sure you commit this often with meaningful messages. 


### Getting Started


We will work with the dataset called [gapminder](https://github.com/jennybc/gapminder), this is a cleaned up version from [Gapminder Data](http://www.gapminder.org/data/). Gapminder contains a lot of great data on all of the nations of the world. We first need to install the gapminder package in R. 

```{r}
install.packages("gapminder")
```



```{r}
library(dplyr)
library(gapminder)
gapminder
```



### Pre-Class Problems

Use **dplyr** functions to address the following questions:

1. How many unique countries are represented per continent?
```{r}
gapminder %>%
group_by(continent) %>%
summarise(countriespercontinent = n_distinct(country))
```
<b> There are 52 countries in Africa, 25 countries in the Americas, 33 countries in Asia, 30 countries in Europe, and 2 countries in Oceania. </b>

2. Which European nation had the lowest GDP per capita in 1997? 

```{r}
gapminder %>%
filter(continent == 'Europe' & year == 1997) %>%
arrange(gdpPercap)
```
<b>Albania had the lowest GDP per capita at 3193.055 </b>

3. According to the data available, what was the average life expectancy across each continent in the 1980s?

```{r}
gapminder %>%
  filter(year == 1980:1989) %>%
  group_by(continent) %>%
  summarise(avglife = mean(lifeExp))
```
<b> The mean life expectancy was:

Africa	56.00775

Americas	67.30333	

Asia	62.84146		

Europe	74.11400

</b>

4. What 5 countries have the highest total GDP over all years combined?

```{r}
gapminder %>%
mutate(gdp = pop*gdpPercap) %>%
arrange(desc(gdp)) %>% 
distinct(country) 
```

<b>The five countries with the highest GDP over all years were 
1) United States				
2) China				
3) Japan				
4) India				
5) Germany </b>

5. What countries and years had life expectancies of _at least_ 80 years? _N.b. only output the columns of interest: country, life expectancy and year (in that order)._

```{r}
gapminder %>%
filter(lifeExp >= 80) %>%
select(country, lifeExp, year)
```
<b> Life expectancy was higher than 80 in the following 22 countries:


country
<fctr>
lifeExp
<dbl>
year
<int>
1) Australia	80.370	2002		
2) Australia	81.235	2007		
3) Canada	80.653	2007		
4) France	80.657	2007	
5) Hong Kong, China	80.000	1997		
6) Hong Kong, China	81.495	2002		
7) Hong Kong, China	82.208	2007		
8) Iceland	80.500	2002		
9) Iceland	81.757	2007		
10) Israel	80.745	2007		
11) Italy	80.240	2002		
12) Italy	80.546	2007		
13) Japan	80.690	1997		
14) Japan	82.000	2002		
15) Japan	82.603	2007		
16) New Zealand	80.204	2007		
17) Norway	80.196	2007		
18) Spain	80.941	2007		
19) Sweden	80.040	2002		
20) Sweden	80.884	2007		
21) Switzerland	80.620	2002		
22) Switzerland	81.701	2007		

</b>

6. What 10 countries have the strongest correlation (in either direction) between life expectancy and per capita GDP?

```{r}
gapminder %>%
group_by(country) %>%
summarise(cor = cor(lifeExp, gdpPercap)) %>%
arrange(desc(abs(cor))) %>%
head(10)
```
<b> 
Lowest


country
<fctr>
cor
<dbl>
Botswana	0.005597046			
Mozambique	0.012169063			
Togo	0.021553585			
Afghanistan	-0.047547003			
Ghana	0.058562873			
Lesotho	0.110846981			
Rwanda	0.133197350			
Libya	0.158127598			
Sierra Leone	-0.191221570			
Uganda	0.219015941	

Highest


country
<fctr>
cor
<dbl>
France	0.9962239			
Austria	0.9929642			
Belgium	0.9927496			
Norway	0.9921416			
Oman	0.9907526			
United Kingdom	0.9898930			
Italy	0.9897600			
Israel	0.9884894			
Denmark	0.9870896			
Australia	0.9864457	
</b>


7. Which combinations of continent (besides Asia) and year have the highest average population across all countries? _N.b. your output should include all results sorted by highest average population_. With what you already know, this one may stump you. See [this Q&A](http://stackoverflow.com/q/27207963/654296) for how to `ungroup` before `arrange`ing. This also [behaves differently in more recent versions of dplyr](https://github.com/hadley/dplyr/releases/tag/v0.5.0).

```{r}
gapminder %>%
filter(continent != "Asia") %>%
group_by(year, continent) %>%
summarise(averagepopulation = mean(pop)) %>%
arrange(desc(averagepopulation)) 
```
<b> 

Countries:

year
<int>
continent
<fctr>
averagepopulation
<dbl>
2007	Americas	35954847		
2002	Americas	33990910		
1997	Americas	31876016		
1992	Americas	29570964		
1987	Americas	27310159		
1982	Americas	25211637		
1977	Americas	23122708		
1972	Americas	21175368		
2007	Europe	19536618	
...
</b>


8. Which three countries have had the most consistent population estimates (i.e. lowest standard deviation) across the years of available data? 
9. Subset **gm** to only include observations from 1992 and store the results as **gm1992**. What kind of object is this?
10. Which observations indicate that the population of a country has *decreased* from the previous year **and** the life expectancy has *increased* from the previous year? See [the vignette on window functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html).
